Discovering causal factors with back-to-back regression

### 2019-05-22 paper/paper.tex

Replacing in the expression of the minimal norm solution yields
\begin{equation}
\begin{aligned}
X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
&=X(K'X)'((K'X)(K'X)')^{-1}((FK)(FK)')^{-1}(FK)' \\
&=(XX')K(K'XX'K)^{-1}(FKK'F')^{-1} K'F'
\end{aligned}
\end{equation}

Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)=S$

Thus we have $X(FEX)^\dagger =S (FKK'F')^{-1} K'F'$

FC: this explains the negative diagonal elements, when X is negatively correlated. Need now to prove that this is S(FE)+

FC: onward to the reduction, I assume that we get a scaled value of S(FE)+, this needs a better proof

If matrices X'X and (FN)'(FN) have low condition numbers, and denoting by VarX and VarN the variance of X and FN, we have $$\min_G \Tr ((I-GFE)'(I-GFE)XX') + \Tr (G'GFNN'F') \approx \left \| (I-GFE)\right\| ^2 VarX  + \left \| G\right \| ^2 VarN$$and the solution has the form $k(FE)^\dagger$, k a positive constant below one. To determine an approximate value of k, note that if r is the rank of FE (the number of ones in E) $$\left \| (I-GFE)\right\| ^2 = r (1-k)^2 + dx - r$$ and that under the hypothesis of F having a low condition number, the squared norm of $k(FE)^\dagger$ should be close to $k^2r$. We are therefore minimising a quantity close to $$ VarX (r (1-k)^2 + dx - r) + VarN k^2 r$$ Zeroing the derivative in k yields an approximate value of k, as $$k= \frac {VarX}{VarX + VarN}=\frac{1}{1+nsr}$$ nsr the noise to signal ratio, VarN divided by VarX.

@@

\subsection{Stability analysis}

>>>>

Replacing in the expression of the minimal norm solution yields
\begin{equation}
\begin{aligned}
X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F' 
\end{aligned}
\end{equation}

Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').

Let us also note that $FE=FKK'$, and so $(K'F'FK)^{-1} K'F'FKK'=I_{k}K'$. Therefore, we have $(K'F'FK)^{-1} K'F'=(FE)^\dagger$.

Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FE)^\dagger$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (that is, if causal features do not act as confounders for inactive features), then S=E, and we are retrieving $E (FE)^\dagger$.

Under centred noise, and if (FN)'(FN) have low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover of the minimum of the left norm, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below 1.

Replacing, we have $\left \| X-GFEX\right\| ^2 = \left \| X-\lambda S (FE)^\dagger FEX\right\| = \left \| X(I-\lambda S)\right\|$. If the covariance of X is block diagonal (no confounding between active and inactive features of X), this is equal to $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ (denoting the variance of X over active and inactive features).

For the right part of the sum, we have $GFN=\lambda S (FE)^\dagger FN= \lambda S (FE)^\dagger (FE+F(I-E))N= \lambda S (E +(FE)^\dagger F(I-E))N=\lambda E N $ (FC last part of proof to be done, need some lemma on (FE)+). The square norm is therefore $\lambda^2 Var(N) (k/nx)$.

The total quantity to be minimised, under noise, but with non confounding features in X, F full rank over E, and FN decently conditioned, is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N) (k/nx)$, setting the partial derivative in $\lambda$ to zero, we obtain $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N) (k/nx)}$, and if we suppose that X has comparable variance over all features, this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).

The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} E (FE)^\dagger$ if active and inactive features of X are not correlated

@@
